---
title: "Meet R"
author: "Otho Mantegazza"
execute: 
  echo: true
format:
  revealjs:
    theme: [simple, style/theme.scss]
    incremental: true
    footer: '2022 CBSER Summer School'
editor_options: 
  chunk_output_type: console
---

# Hands On Now 👐   

## R understands you

Write someting at the R console:

. . .

A simple number...
 
```{r}
2
```

. . .

Some operation...

```{r}
3 + 7
```

. . .

Text must be quoted...

```{r}
'hello'
```

. . .

Otherwise it is interpreted as the name of a variable:

```{r}
#| error: true
my_var
```

## Variables

You can define a variable with the operator `<-` .

```{r}
# a numeric variable
this_year <- 2022
 
# a character variable
italian_greetings <- 'ciao'
```

. . .

If you type the name of a variable without quotes, r prints its value:

```{r}
this_year
```

```{r}
italian_greetings
```

. . .

Use the keyboard shortcut `option` + `-` to write the assign operator `<-` more easily.

## Data Types

When you define an object R guesses its type.

. . .

`typeof()` returns the internal type of an object.

```{r}
typeof(this_year)
typeof(italian_greetings)
```

. . .

`class()` returns a the class attribute of an object.

```{r}
class(this_year)
class(italian_greetings)
```

## Everything in R is an object

 and has a type.


