---
title: "Meet R"
author: "Otho Mantegazza"
execute: 
  echo: true
format:
  revealjs:
    theme: [simple, style/theme.scss]
    incremental: true
    footer: '2022 CBSER Summer School'
editor_options: 
  chunk_output_type: console
---

# Hands On Now 👐   

## R understands you

Write someting at the R console:

. . .

A simple number...
 
```{r}
2
```

. . .

Some operation...

```{r}
3 + 7
```

. . .

Text must be quoted...

```{r}
'hello'
```

. . .

Otherwise it is interpreted as the name of a variable:

```{r}
#| error: true
my_var
```

## You can define variables

You can define a variable with the operator ` <- ` .

```{r}
# a numeric variable
this_year <- 2022
 
# a character variable
italian_greetings <- 'ciao'
```

. . .

If you type the name of a variable without quotes, r prints its value:

```{r}
this_year
```

```{r}
italian_greetings
```

. . .

Use the keyboard shortcut ` alt/option ` + ` - ` to write the assign operator ` <- ` more easily.

## Data Types

When you define an object R guesses its type.

. . .

`typeof()` returns the internal type of an object.

```{r}
typeof(this_year)
typeof(italian_greetings)
```

. . .

`class()` returns a the class attribute of an object.

```{r}
class(this_year)
class(italian_greetings)
```

## Everything in R is an object

 And objects have a types.

. . .

```{r}
class(2)
```

. . .

```{r}
class('hello')
```

. . .

```{r}
class(FALSE)
```

. . .

```{r}
class(mean)
```

. . .

```{r}
class(`<-`)
```

. . .

```{r}
class(Sys.Date())
```

## Exercise {.exercise}

1. Use the function `c(arg_1, arg_2, ..., arg_n)` to put together in a vector many objects that you expect to have the same class. Try with different ones.

2. Now do the same, but try to put together in a vector data with different types, what happened?

::: aside
Learn more about [vectors and types at Advanced R](https://adv-r.hadley.nz/vectors-chap.html).
:::

# All Data Are Vectors

## Vectors store one or more data point of the same class

```{r}
num_vector <- c(1, 3, 5, 3, 6)

character_vector <- c('Hi', ',', 'I', 'am', 'a', 'character', 'vector')

logical_vector <- c(TRUE, FALSE, FALSE, TRUE, FALSE)
```

. . .

<br>

```{r}
class(num_vector)
class(character_vector)
class(logical_vector)
```

## Classes can be coherced

`TRUE` and `FALSE` can be coerced into numeric zeroes and ones, or also into characters.

```{r}
c(2, TRUE, 5, FALSE, 4)
```

. . .

Numbers can be coerced into characters.

```{r}
c('a', TRUE, 5, FALSE, 4)
```

## Classes can be coherced

Classes can be coerced explicitly.

```{r}
logical_vector <- c(T, F, F, T, T, F)
logical_vector
as.numeric(logical_vector)
as.character(logical_vector)
```


. . .

If an element can't be coerced to the desired class, R returns a missing value (`NA`).

```{r}
as.numeric(c('1', 'two', '3'))
```


## Atomic elements are vectors of length 1

```{r}
short_vector <- 1
long_vector <- 1:50
```

. . .



```{r}
short_vector
long_vector
```

. . .

```{r}
class(short_vector)
class(long_vector)
```

. . .

```{r}
length(short_vector)
length(long_vector)
```

## Most functions in R are vectorized

If I want to apply an operation over a vector, I can just write it as it is without a for loop. For example:

. . . 

Add an integer to every number in a vector:

```{r}
long_vector + 100
```

. . .

Multiply every number in a vector by an integer:

```{r}
long_vector * 2
```

## Most functions in R are vectorized

Take the square root of every number in a vector:

```{r}
sqrt(long_vector)
```

## Exercise {.exercise}

1. Generate a vector of 100 (semi-) random numbers with a normal distribution.

2. Use a for loop (even if you don't need one) to multiply eachs number in the vector by 2.

:::aside
[For loops in R](https://bookdown.org/rdpeng/rprogdatascience/control-structures.html#for-loops) according to Roger D. Peng.
:::

## Extra - Lists

If you need to collect together and store data of different types, you can use lists.

```{r}
my_list <- list(
  'hello',
  1,
  FALSE,
  1:25
)

my_list
```


# Let's Build a Dataframe 🧱

## Can we use vectors to build a dataframe? 

We learned that vector contain elements of the same type, for example, numerics, character, logical.

. . .

According to the **[Tidy Data](https://tidyr.tidyverse.org/articles/tidy-data.html)** theory, in rectangular dataframes:

::: {.nonincremental}

- Each row is an observation.
- Each column is a variable.

:::

By definiton variables of your data, must be made by atomic elements of the same type. So we can use vectors of the same length to build the columns of a dataframe.

Let's try.

## And... We have data 

Let's prepare three vectors with the first observations of `iris` and use them to make a dataframe.

```{r}
sepal_length <- c(5.1, 7.0, 6.3)

sepal_width <- c(3.5, 3.2, 3.3)

species <- c('setosa', 'versicolor', 'virginica')

iris_simple <- 
  data.frame(
    species,
    sepal_length,
    sepal_width
  )

iris_simple
```

## And... We have data 

The variable `iris_simple` points to a dataframe that stores rectangular data.

```{r}
class(iris_simple)
```

```{r}
dim(iris_simple)
```

. . .

We can access its columns with the operator `$`. Each column is a vector.

```{r}
iris_simple$sepal_length
```

```{r}
class(iris_simple$sepal_length)
```

```{r}
is.vector(iris_simple$sepal_length)
```

## {background-image='img/stephan-quintet.jpg'}

[Now that we have data, let's take our first stroll into the Tidyverse.]{style="color:#ffffff; font-size:80px"}

## Tibbles: a modern take on dataframes

Let's attach the whole [Tidyverse](https://www.tidyverse.org/), which includes also the [Tibble](https://tibble.tidyverse.org/) package...

```{r}
library(tidyverse)
```

. . .

...and convert our dataframe `iris_simple` into a tibble.

```{r}
iris_simple <- as_tibble(iris_simple)
```

. . .

The print method for a `tibble`, provides a handy and informative output.

```{r}
iris_simple
```

## Tibbles: a modern take on dataframes {auto-animate="true"}

Let's compare the print method for the dataframe `iris`, before and after we convert it to a tibble.

```{r}
iris
```

## Tibbles: a modern take on dataframes {auto-animate="true"}

Let's compare the print method for the dataframe `iris`, before and after we convert it to a tibble.

```{r}
as_tibble(iris)
```

## Exercise {.exercise}

::: {.nonincremental}

1. Check the weather forecast for your hometown (or a place of your choice).

2. Assign the forecasted temperature in one vector and the corresponding time in another vector for at least 5 data points.

3. Make a dataframe with those two variables.

4. What's the average temperature? And what's its standard deviation?

5. Then, visualize those data with a line-chart in which the x axis represents time and the y axis represents the forecasted temperature.

:::



