---
title: "Manipulate Data"
subtitle: "Tidyverse - Part 1"
author: "Otho Mantegazza"
execute: 
  echo: true
format:
  revealjs:
    theme: [simple, style/theme.scss]
    incremental: true
    footer: '[2022 CBSER Summer School](https://www.summercompschool.mawazoinstitute.org/)'
editor_options: 
  chunk_output_type: console
---

```{r}
#| include: false
library(dplyr)
library(magrittr)
library(tidyr)
library(palmerpenguins)
library(skimr)
```


# Intro to the Tidyverse

## The [Tidyverse](https://www.tidyverse.org/packages/) is an ecosystem of packages for Data Science

All the packages share a common design:

  - One function does one thing, well.
  - Designed for pipes.
  - Extensive documentation.
  - Non-standard evaluation, to write code quickly and easily.

. . .
  
All packages can be loaded with `library(tidyverse)`, but you can also load single packages one by one.

# Data: Palmer Penguins {background-image="https://upload.wikimedia.org/wikipedia/commons/a/a3/Gentoo_Penguin_AdF.jpg"}

:::small
[Photo credits: Arturo de Frias Marques]
:::

## {background-image="https://allisonhorst.github.io/palmerpenguins/reference/figures/lter_penguins.png" background-position='center' background-size='contain'}

## We first explore the Palmer Penguins Dataset

The [Penguins Dataset](https://allisonhorst.github.io/palmerpenguins/) stores real data about palmer penguins. This R data package was developed and is maintained by Allison Horst, Alison Hill and Kirsten Gorman for teaching purposes.

. . . 

Let's install the package...

```{r}
#| eval: false
install.packages('palmerpenguins')
```

. . . 

...and load it in R.

```{r}
library(palmerpenguins)
```

## `palmerpenguins` exports two datasets {auto-animate="true"}

```{r}
penguins_raw
```


## `palmerpenguins` exports two datasets {auto-animate="true"}

```{r}
penguins
```

We will use the first one: `penguins`, which has already been cleaned.

## Exercise {.exercise}

The print method for a tibble gives you a reasonable overview of the data stored in it.

Can you get more details with the package [skimr](https://docs.ropensci.org/skimr/)?

Check its documentation, install it, try it out on the `penguins` dataset. Comment on the output. 


# Tool: dplyr

## A grammar for data manipulation

:::: {.columns}

::: {.column width="40%"}
![](https://raw.githubusercontent.com/rstudio/hex-stickers/master/SVG/dplyr.svg)
:::

::: {.column width="60%" .smallish}

[Dplyr](https://dplyr.tidyverse.org/) provides a grammar for manipulating data, with many useful verbs:

- `mutate()` adds new variables that are functions of existing variables
- `select()` picks variables based on their names.
- `filter()` picks cases based on their values.
- `summarise()` reduces multiple values down to a single summary.
- `group_by()` performs operations by group.

:::

::::

## We can apply those verbs to manipulate data {auto-animate="true"}

```{r}
penguins %>%
  mutate(bill_length_meters = bill_length_mm/1000)
```


## We can apply those verbs to manipulate data {auto-animate="true"}

```{r}
penguins %>%
  select(bill_length_mm) %>% 
  mutate(bill_length_meters = bill_length_mm/1000)
```

## We can apply those verbs to manipulate data {auto-animate="true"}

```{r}
penguins %>%
  select(species, island, bill_length_mm) %>% 
  mutate(bill_length_meters = bill_length_mm/1000)
```

## We can apply those verbs to manipulate data {auto-animate="true"}

```{r}
penguins %>%
 count(island)
```

## We can apply those verbs to manipulate data {auto-animate="true"}

```{r}
penguins %>%
  select(species, island, bill_length_mm) %>% 
  filter(island == 'Dream') %>% 
  mutate(bill_length_meters = bill_length_mm/1000)
```

## We can apply those verbs to manipulate data {auto-animate="true"}

```{r}
penguins %>%
  select(species, island, bill_length_mm) %>% 
  filter(island == 'Dream') %>% 
  mutate(bill_length_meters = bill_length_mm/1000) %>% 
  group_by(species)
```

## We can apply those verbs to manipulate data {auto-animate="true"}

```{r}
penguins %>%
  select(species, island, bill_length_mm) %>% 
  filter(island == 'Dream') %>% 
  mutate(bill_length_meters = bill_length_mm/1000) %>% 
  group_by(species) %>% 
  summarise(mean_bill_length_mm = mean(bill_length_mm),
            sd_bill_length_mm = sd(bill_length_mm))
```


## We can apply those verbs to manipulate data {auto-animate="true"}

```{r}
dream_summary <- 
  penguins %>%
  select(species, island, bill_length_mm) %>% 
  filter(island == 'Dream') %>% 
  mutate(bill_length_meters = bill_length_mm/1000) %>% 
  group_by(species) %>% 
  summarise(mean_bill_length_mm = mean(bill_length_mm),
            sd_bill_length_mm = sd(bill_length_mm))
```

Let's assign the output to a new variable `dream_summary`.

## {.big}

In the previous code we have seen also two additional aspects that feature heavily in the tidyverse:

- The Pipe `%>%`.
- Non-Standard Evaluation.

## The Pipe `%>%`

:::: {.columns}

::: {.column width="40%"}
![](https://raw.githubusercontent.com/rstudio/hex-stickers/master/SVG/pipe.svg)
:::

::: {.column width="10%"}
:::

::: {.column width="50%"}

The pipe is provided by the package [magrittr](https://magrittr.tidyverse.org/), it's a forwarding operator:

<br>

It takes the ouput of what comes before (LHS) and sends it to the first argument of the function that comes after (RHS).

<br>

[`LHS %>% RHS`]{.big}

:::

::::

## The Pipe `%>%` {auto-animate="true"}

For example, you could write:

```{r}
select(penguins, species, body_mass_g)
```

## The Pipe `%>%` {auto-animate="true"}

...but if you use the pipe, your code is easier to read...

```{r}
penguins %>% select(species, body_mass_g)
```

## The Pipe `%>%` {auto-animate="true"}

...especially if you have to perform many operations one after the other...

```{r}
penguins %>%
  select(species, body_mass_g) %>% 
  filter(species == 'Adelie')
  
```

## The Pipe `%>%` {auto-animate="true"}

...that otherwise would force you to nest your code horribly.

```{r}
filter(
  select(
    penguins,
    species, body_mass_g
  ),
  species == 'Adelie'
)

```

## Non Standard Evaluation

This one is hard

