---
title: "Manipulate Data"
subtitle: "Tidyverse - Part 1"
author: "Otho Mantegazza"
execute: 
  echo: true
format:
  revealjs:
    theme: [simple, style/theme.scss]
    incremental: true
    footer: '[2022 CBSER Summer School](https://www.summercompschool.mawazoinstitute.org/)'
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
library(dplyr)
library(magrittr)
library(tidyr)
library(palmerpenguins)
library(skimr)
```


# Intro to the Tidyverse

## The [Tidyverse](https://www.tidyverse.org/packages/) is an ecosystem of packages for Data Science

All the packages share a common design:

  - One function does one thing, well.
  - Designed for pipes.
  - Extensive documentation.
  - Non-standard evaluation, to write code quickly and easily.

. . .
  
All packages can be loaded with `library(tidyverse)`, but you can also load single packages one by one.

# {background-image="https://allisonhorst.github.io/palmerpenguins/reference/figures/lter_penguins.png" background-position='center' background-size='contain'}

## We first explore the Palmer Penguins Dataset

The [Penguins Dataset](https://allisonhorst.github.io/palmerpenguins/) stores real data about palmer penguins. This R data package was developed and is maintained by Allison Horst, Alison Hill and Kirsten Gorman for teaching purposes.

. . . 

Let's install the package...

```{r}
#| eval: false
install.packages('palmerpenguins')
```

. . . 

...and load it in R.

```{r}
library(palmerpenguins)
```

## `palmerpenguins` exports two datasets {auto-animate="true"}

```{r}
penguins_raw
```


## `palmerpenguins` exports two datasets {auto-animate="true"}

```{r}
penguins
```

We will use the first one: `penguins`, which has already been cleaned.