---
title: "Notes on Correlation"
author: "Otho Mantegazza"
execute: 
  echo: true
format:
  html:
    theme: [style/page.scss]
    toc: true
    number-sections: true
    footer:  '[Home](../index.html) | [2022 CBSER Summer School](https://www.summercompschool.mawazoinstitute.org/)'
editor_options: 
  chunk_output_type: console
---

# Generic intro

For a generic intro to correlation check the links below:

- [Making Sense of Data With R - Correlation](https://bookdown.org/yshang/book/correlation.html)
- [Toward Data Science - Pearson vs Spearman Correlation Coefficient](https://towardsdatascience.com/clearly-explained-pearson-v-s-spearman-correlation-coefficient-ada2f473b8)

# Correlation in R

You can estimate the Pearson or Spearman correlation coefficient between two numerical vectors or matrices with the function [`cor()`](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/cor).

# Example on  Palmer Penguins

Let's check an example on the Penguins dataset:

```{r, message=FALSE}
library(tidyverse)
library(palmerpenguins)
library(corrplot)
library(GGally)
```

Remember the [palmer penguins](https://allisonhorst.github.io/palmerpenguins/) dataset?

```{r}
penguins
```

## Visualize first

Let's say that i would like to measure the correlation between the two variables `flipper_length_mm` and `body_mass_g`.

Let's visualize them in a scatterplot first:

```{r}
penguins %>% 
  ggplot() +
  aes(x = flipper_length_mm,
      y = body_mass_g,
      colour = species) +
  geom_point()
```

Il looks like there's strong overall correlation between the two variables, which might be weaker in the Adelie species.

## Estimate the correlation coefficient

Let's estimate the Pearson correlation coefficient with the function `cor()`.

```{r}
cor(
  x = penguins$flipper_length_mm,
  y = penguins$body_mass_g,
  use = 'pairwise.complete.obs',
  method = 'pearson'
)
```

We use the argument `use = 'pairwise.complete.obs'` to remove observation that have missing values in x or y.

The two variables are highly correlated, with a score of 0.87.

## Stratify by species

We can also use the group_by-summarize paradigm to estimate the same Pearson correlation coefficient, together with the Spearman correlation coefficient, stratified by species.

```{r}
penguins %>% 
  group_by(species) %>% 
  summarise(
    pearson_corr = cor(
      x = flipper_length_mm,
      y = body_mass_g,
      use = 'pairwise.complete.obs',
      method = 'pearson'
    ),
    spearman_corr = cor(
      x = flipper_length_mm,
      y = body_mass_g,
      use = 'pairwise.complete.obs',
      method = 'spearman'
    )
  )
```

The coefficient separated by species are smaller than the one taking all the penguins together. Probably because the trend between body mass and flipper length is conserved across species on a wide range of body mass.

## Correlation matrices

You can use the function `cor()` to measure the correlation of multiple variable at once, producing a correlation matrix.

For example, if we want to check how each continuous variable in the Palmer Penguins dataset correlates with each other:

```{r}
cor_mat <- 
  penguins %>%
  select_if(is.numeric) %>% 
  # let's remove the 'year' variable
  select(-year) %>% 
  cor(use = 'pairwise.complete.obs')

cor_mat
```


# Useful visual packages

R has a couple (or more) useful packages that let you explore how multiple variables correlate across each other.

## corrplot

[corrplot](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html) let's you explore visually correlation matrices.

For example, let's visualize the correlation matrix that we have stored in the variable `cor_mat` in the [section above](#correlation-matrices).

```{r}
cor_mat %>% corrplot(diag = FALSE)
```

```{r}
cor_mat %>% corrplot(method = 'number', diag = FALSE)
```

Don't take all these correlation values for granted. As we learn in the section [caveats on correlation](#caveats-on-correlation), they might hide some spurious correlation value.

## Pair Plot with GGally

Let's use the `ggpairs()` function from the package [GGally](https://ggobi.github.io/ggally/reference/ggpairs.html) to explore at the same time the scatterplots and the correlation coefficient.

```{r}
#| message: false

penguins %>%
  select_if(is.numeric) %>% 
  # let's remove the 'year' variable
  select(-year) %>% 
  # let's drop na's
  drop_na() %>% 
  ggpairs()
```

Find more examples on the [R Graph Gallery](https://r-graph-gallery.com/199-correlation-matrix-with-ggally.html).

# Caveats on Correlation

Correlation is a descriptive statistics. When you measure it, you reduce a complex bivariate distribution to a single number. This reductive approach is generally useful, but might be misleading in several cases .

The main caveats to keep in mind are:

- **Visualize the distribution**. In many specific cases, for example when there are outliers in the dataset, a high correlation scores does not translate into any evident association between the two variables.
- **Correlation does not mean causation**. Even if two variable strongly correlate, this is no evidence that changes in one variable is causing an effect on the other. Correlation alone might be a hint to causation, but it is in no way at all a proof of it.

⚠️⚠️⚠️ For a detailed explanation of the caveats of correlation, check [Chapter 19 of the book Introduction to Statistics with R](http://rafalab.dfci.harvard.edu/dsbook/association-is-not-causation.html) by Rafael A. Irizarry 🙏🙏🙏.

# Exercise

`bill_depth_mm` and `bill_length_mm` correlate negatively with a Pearson correlation coefficient of `-0.235`. Is this correlation true or spurious? Why?

Can you find a stratifying variable that solves this issue?