---
title: "Robust Summary Statistics"
author: "Otho Mantegazza"
execute: 
  echo: true
format:
  revealjs:
    theme: [simple, style/theme.scss]
    footer: '[Home](../index.html) | [2022 CBSER Summer School](https://www.summercompschool.mawazoinstitute.org/)'
editor_options: 
  chunk_output_type: console
---

```{r}
#| include: false
library(dplyr)
library(readr)
library(magrittr)
library(tidyr)
library(palmerpenguins)
library(tibble)
library(ggplot2)
library(readr)
library(here)
library(ggbeeswarm)

theme_update(axis.title = element_text(size = 15),
             axis.text = element_text(size = 15),
             strip.text = element_text(size = 15))

violet <- '#6d07a8'
```

# Summary Statistics

## Concise summaries

When we mentioned the boxplot, we talked about robust summary statistics. First of all, what is a summary statistic?

A Summary statistic describes a property of a distribution in one single number.

For example:

- The **mean** represents the **central tendency** of a distribution.
- The **standard deviation** represents its dispersion.

**None** of these statistics is robust toward outliers.

## Mean and Standard Deviation {.scrollable}

### - Mean

```{r}
#| echo: false
penguins_summary <- function(pengs) {
  pengs %>% 
    group_by(species) %>% 
    summarise(m = mean(bill_depth_mm, na.rm = T),
              sd = sd(bill_depth_mm, na.rm = T),
              med = median(bill_depth_mm, na.rm = T),
              iqr = IQR(bill_depth_mm, na.rm = T),
              mad = mad(bill_depth_mm, na.rm = T))
}

penguins_summarized <- 
  penguins %>% 
  penguins_summary()

p <- 
  penguins %>% 
  ggplot(aes(x = bill_depth_mm,
             fill = species)) +
  geom_histogram(binwidth = .4) +
  labs(y = 'count') +
  guides(fill = 'none') +
  facet_grid(rows = 'species')
```


::: {style="color:red;"}

$mean(x) = \frac{1}{n} \sum_{i=i}^{n} x_{i}$ 

:::

::: {style="height: 750px; width: 1000px;"}
 
```{r}
#| echo: false
#| fig.height: 7
#| fig-width: 7

p_mean <- 
  p +
  geom_vline(data = . %>% penguins_summary(),
            aes(xintercept = m),
            colour = 'red',
            size = 2,
            lty = '11') +
  geom_label(data = . %>% penguins_summary(),
            aes(x = m,
                y = 1,
                label = m %>% round(1)),
            fill = 'white',
            colour = 'red',
            hjust = -0.2,
            vjust = 0)

p_mean
```

:::

### - Standatrd Deviation

::: {style="color:red;"}

$SD = \sqrt{\frac{1}{N-1} \sum_{i=1}^N \left(x_i - \bar{x}\right)^2}$

:::
 
::: {style="height: 750px; width: 1000px;"}
 
```{r}
#| echo: false
#| fig.height: 7
#| fig-width: 7

p_sd <- 
  p +
  geom_segment(data = . %>% 
              group_by(species) %>% 
              summarise(m = mean(bill_depth_mm, na.rm = T),
                        sd = sd(bill_depth_mm, na.rm = T)),
            aes(x = m-sd/2,
                xend = m+sd/2,
                y = 15,
                yend = 15),
            lty = '11',
            colour = 'red',
            size = 2) +
  geom_label(data = . %>% penguins_summary(),
            aes(x = m,
                y = 15,
                label = sd %>% round(1)),
            fill = 'white',
            colour = 'red',
            hjust = .5,
            vjust = -.4) 
p_sd
```

:::

## Mean and SD are not robust to outlier

What if I'm distracted and forget to write a comma when I collect the data.

```{r}
penguins$bill_depth_mm %>% range(na.rm = T)
```

With this snipped we multiply the max value of `bill_depth_mm` by 10.

```{r}
penguins_bad <- 
  penguins %>% 
    mutate(
      bill_depth_mm = bill_depth_mm %>% 
        {
          if_else(. == max(., na.rm = T),
                    true = .*10,
                    false = .)
        }
    )
```

Let's see what happens to the mean and to the SD.

## {.scrollable}
::: {style="height: 750px; width: 1000px;"}
 
```{r}
#| echo: false
#| fig.height: 7
#| fig-width: 7
penguins_summary <- function(pengs) {
  pengs %>% 
    group_by(species) %>% 
    summarise(m = mean(bill_depth_mm, na.rm = T),
              sd = sd(bill_depth_mm, na.rm = T),
              med = median(bill_depth_mm, na.rm = T),
              mad = mad(bill_depth_mm, na.rm = T))
}

penguins_bad %>% 
  ggplot(aes(x = bill_depth_mm,
             fill = species)) +
  geom_histogram(binwidth = .4) +
  geom_vline(data = . %>% penguins_summary(),
            aes(xintercept = m),
            colour = 'red',
            size = 2,
            lty = '11') +
  geom_label(data = . %>% penguins_summary(),
            aes(x = m,
                y = 1,
                label = m %>% round(1)),
            fill = 'white',
            colour = 'red',
            hjust = -0.2,
            vjust = 0) +
  guides(fill = 'none') +
  facet_grid(rows = 'species')
```

:::

### - Standard Deviation

::: {style="color:#6d07a8;"}

$SD(x) = \sqrt{\frac{1}{N-1} \sum_{i=1}^N \left(x_i - \bar{x}\right)^2}$

:::
 
::: {style="height: 750px; width: 1000px;"}
 
```{r}
#| echo: false
#| fig.height: 7
#| fig-width: 7
penguins_bad %>% 
  ggplot(aes(x = bill_depth_mm,
             fill = species)) +
  geom_histogram(binwidth = .4) +
  geom_segment(data = . %>% 
              group_by(species) %>% 
              summarise(m = mean(bill_depth_mm, na.rm = T),
                        sd = sd(bill_depth_mm, na.rm = T)),
            aes(x = m-sd/2,
                xend = m+sd/2,
                y = 15,
                yend = 15),
            lty = '11',
            colour = '#6d07a8',
            size = 2) +
  geom_label(data = . %>% penguins_summary(),
            aes(x = m,
                y = 15,
                label = sd %>% round(1)),
            fill = 'white',
            colour = '#6d07a8',
            hjust = .5,
            vjust = -.4) +
  guides(fill = 'none') +
  facet_grid(rows = 'species')
```

:::

## Reference

https://rafalab.github.io/dsbook/robust-summaries.html